---
title: 'Analysis of Clustered and Longitudinal Data: Project 1'
author: "Runa Broekaert, Kobe De Neve and Bart Smets"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(nlme)
library(gridExtra)
library(corrplot)
```

## Load and inspect data

```{r}
dat <- read.table("income1968.dat", header = TRUE, stringsAsFactors = TRUE)
dat <- within(dat, person <- factor(person))

str(dat)
summary(dat)
```

## Plots

```{r}
ggplot(dat, aes(x = year, y = income, group = person, col = sex)) +
  geom_line(alpha = 0.6) +
  facet_wrap(facet = ~ sex) +
  theme_minimal() +
  guides(col = "none")
```

What do we see?

-   Globally:

    -   some imbalance/missing data, because not all subjects were measured every year, e.g. subject 11 has no measurement for 1968, subject 12 has no measurement for 1968-1969 and 1988-1990, ...

    -   increasing trend

    -   individual variability

    -   'fanning out': smaller variance at baseline, more variation towards the end \> suggests random intercepts and slopes

-   Gender differences:

    -   more variation and more outliers for men, also confirmed by boxplots for income per gender:

```{r}
ggplot(dat, aes(x = sex, y = income, col = sex)) +
  geom_boxplot() +
  theme_minimal() +
  guides(col = "none")
```

The mean structure looks as follows:

```{r, message = FALSE}
ggplot(dat, aes(x = year, y = income, col = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  facet_wrap(facet = ~ sex) +
  theme_minimal() +
  guides(col = "none")
```

What do we see?

-   smoothers are relatively linear

-   variance of data points around smoother increases with time

Why might log-transform of income be useful?

```{r, message = FALSE}
ggplot(dat, aes(x = year, y = log(income), group = person, col = sex)) +
  geom_line(alpha = 0.6) +
  facet_wrap(facet = ~ sex) +
  theme_minimal() +
  guides(col = "none")

ggplot(dat, aes(x = year, y = log(income), col = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  facet_wrap(facet = ~ sex) +
  theme_minimal() +
  guides(col = "none")

income_hist <- ggplot(dat, aes(x = income)) +
  geom_histogram() +
  theme_minimal()

log_income_hist <- ggplot(dat, aes(x = log(income))) +
  geom_histogram() +
  theme_minimal()

grid.arrange(income_hist, log_income_hist, nrow = 1)
```

What do we see?

-   variance is more constant (no 'fanning' as before)

-   marginal distribution of income is more symmetrical

Next, we assess the relationship between the log-transformed income and the remaining variables.

log(income) against age:

```{r}
ggplot(dat, aes(x = age, y = log(income), col = sex)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  labs(x = "age at baseline") +
  facet_wrap(facet = "sex") +
  theme_minimal() +
  guides(col = "none")

ggplot(dat, aes(x = factor(age), y = log(income), col = sex)) +
  geom_boxplot() +
  labs(x = "age at baseline") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

What do we see?

-   relatively horizontal mean structure, but some non-linearity and increasing trend for men after age 32

-   no very clear variance pattern

log(income) against years of education:

```{r}
ggplot(dat, aes(x = educ, y = log(income), col = sex)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  facet_wrap(facet = ~ sex) +
  labs(x = "years of education") +
  theme_minimal() +
  guides(col = "none")

ggplot(dat, aes(x = factor(educ), y = log(income), col = sex)) +
  geom_boxplot() +
  labs(x = "years of education") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

What do we see?

-   for some levels of years of education, only one of the two genders is represented

-   some none-linearity in mean structure

-   some increase in variance

Finally, we have a look at the correlation between measurements at different time points.

```{r}
dat_wide <- pivot_wider(dat, names_from = "year", values_from = "income")
dat_wide[, 5:27] <- log(dat_wide[, 5:27])

corrplot(cor(dat_wide[, 5:27], use = "na.or.complete"),
         method = "number",
         type = "upper")
```

What do we see?

-   correlations of log(income) at different measurement times

-   correlations decrease over time

-   NOTE: missing values \> is this plot useful??

-   didn't do a scatterplot matrix using ggpairs() because there are too many time points

## Initial model

Considerations for our first model with age and sex:

-   mean structure

    -   start with maximal mean structure, so include interaction
    -   treat year variable as a factor, we will later check whether we can assume linearity

-   covariance structure

    -   plots above showed a relatively homogeneous variance over time (constant variance on diagonal of covariance matrix)

    -   need to take into account clustering of observations within people

    -   since we have longitudinal data, it is likely that observations within subjects closer in time are more correlated than observations that are further apart in time; the correlation plot also showed this, so we need to account for it.

    -   the measurements are annual, i.e. made at approximately equal time intervals

    -\> random intercept + first-order autoregressive

```{r}
dat <- within(dat, log_income <- log(income))
dat <- within(dat, cyear <- factor(year))

fit1 <- lme(
  log_income ~ sex * cyear,
  random = ~ 1 | person,
  correlation = corCAR1(form = ~ year | person),
  weights = varIdent(form = ~ 1),
  data = dat,
  na.action = na.omit,
  method = "REML"
)
```

Do we need autoregressive covariance structure?

```{r}
fit2 <- lme(
  log_income ~ sex * cyear,
  random = ~ 1 | person,
  correlation = NULL,
  weights = NULL,
  data = dat,
  na.action = na.omit,
  method = "REML"
)

anova(fit1, fit2)
```

Checking the mean structure: can we assume linearity for year?

```{r}
fit1 <- lme(
  log_income ~ sex * cyear,
  random = ~ 1 | person,
  correlation = corCAR1(form = ~ year | person),
  weights = varIdent(form = ~ 1),
  data = dat,
  na.action = na.omit,
  method = "ML" # refit first model with ML
)

fit3 <- lme(
  log_income ~ sex * year,
  random = ~ 1 | person,
  correlation = corCAR1(form = ~ year | person),
  weights = NULL,
  data = dat,
  na.action = na.omit,
  method = "ML"
)

anova(fit1, fit3)
```

## Random slopes

Do we need random slopes for year? Both test and AIC say yes:

```{r}
fit1 <- lme(
  log_income ~ sex * year,
  random = ~ 1 | person,
  correlation = corAR1(form = ~ year | person),
  weights = NULL,
  data = dat,
  na.action = na.omit,
  method = "REML"
)

fit5 <- lme(
  log_income ~ sex * year,
  random = ~ year | person,
  correlation = corAR1(form = ~ year | person),
  weights = NULL,
  data = dat,
  na.action = na.omit,
  method = "REML"
)

lrr <- logLik(fit1)
lrf <- logLik(fit5)
lr <- as.numeric(-2 * (lrr - lrf))
df1 <- 1 # right degrees of freedom????
df2 <- 2
p1 <- pchisq(lr, df1, lower.tail = FALSE)
p2 <- pchisq(lr, df2, lower.tail = FALSE)
p <- (p1 + p2) * 0.5
lrt <- data.frame(
  lr = lr,
  df1 = df1,
  df2 = df2,
  p = p
)
lrt
```

```{r}
AIC(fit1, fit5)
```

Interpretation of random slope

```{r}
coef(summary(fit5))
```

```{r}
VarCorr(fit5)
```

For 95% of subjects, the average annual increase in log income lies between `0.09 - 1.96 * 0.039 = 0.0312` and `0.09 + 1.96 * 0.039 = 0.1488`.

## Covariance between random slopes and intercepts

```{r}
blups <- as.data.frame(as.matrix(ranef(fit5)))
blups$Subject <- row.names(blups)

ggplot(blups, aes(x = `(Intercept)`, y = year, label = Subject)) +
  geom_text() + 
  labs(x = "Intercept", y = "Year") +
  theme_minimal()
```

```{r}
fit6 <- lme(
  log_income ~ sex * year,
  random = list(person = pdDiag(form = ~ year)),
  correlation = corCAR1(form = ~ year | person),
  weights = NULL,
  data = dat,
  na.action = na.omit,
  method = "REML"
)

AIC(fit5, fit6)
```

## Form of year effect

Plot of residuals against year to detect non-linearity

```{r}
# linear vs quadratic
fit6_ml <- lme(
  log_income ~ sex * year,
  random = list(person = pdDiag(form = ~ year)),
  correlation = corCAR1(form = ~ year | person),
  data = dat, na.action = na.omit, method = "ML"
)
fit_quad_ml <- lme(
  log_income ~ sex * year + I(year^2),
  random = list(person = pdDiag(form = ~ year)),
  correlation = corCAR1(form = ~ year | person),
  data = dat, na.action = na.omit, method = "ML"
)
anova(fit6_ml, fit_quad_ml)
AIC(fit6_ml, fit_quad_ml)

```

```{r}
# Piecewise linear  (knot at year 75 but don't think it's necessary)
dat <- within(dat, year_75 <- pmax(year - 75, 0))

fit_spl_ml <- lme(
  log_income ~ sex * year + year_75,
  random = list(person = pdDiag(form = ~ year)),
  correlation = corCAR1(form = ~ year | person),
  data = dat, na.action = na.omit, method = "ML"
)

anova(fit6_ml, fit_spl_ml)
AIC(fit6_ml, fit_spl_ml)

```

We find a non-significant p-value twice, implying that both a quadratic trend and a piecewise linear model do not improve on a linear model.

```{r}
#Test if age and educ should be included in the model.
fit_cov_ml <- lme(
  log_income ~ sex * year + age + educ,
  random = list(person = pdDiag(form = ~ year)),
  correlation = corCAR1(form = ~ year | person),
  data = dat, na.action = na.omit, method = "ML"
)

anova(fit6_ml, fit_cov_ml)  


# Results previous analysis significant-> Age and educ interaction with year test
fit_inter_ml <- lme(
  log_income ~ sex * year + age * year + educ * year,
  random = list(person = pdDiag(form = ~ year)),
  correlation = corCAR1(form = ~ year | person),
  data = dat, na.action = na.omit, method = "ML"
)

anova(fit_cov_ml, fit_inter_ml) 

# REML for inference
fit_final <- lme(
  log_income ~ sex * year + age + educ,
  random = list(person = pdDiag(form = ~ year)),
  correlation = corCAR1(form = ~ year | person),
  data = dat, na.action = na.omit, method = "REML"
)

coef(summary(fit_final))
anova(fit_final, type = "marginal")
```
```{r}

dat$pred_log <- predict(fit_final, level = 0)

ggplot(dat, aes(x = year, y = pred_log, colour = sex)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.3) +
  scale_colour_manual(values = c("F" = "coral", "M" = "steelblue"),
                      labels = c("Female", "Male")) +
  labs(x = "year", y = "Pred log(income)",
       title = "Predicted log(income) by sex",
       colour = "Sex") +
  theme_minimal()
```
No interaction necessary as slopes divergent!
